\section{Data}\label{sec:data}

\subsection{ISTAT: Emigration Flows from the Italian Side}
\label{subsec:data_istat}
The Italian side of the comparison is drawn from the \textit{Cancellazioni
anagrafiche per trasferimento di residenza all'estero} dataset, published
annually by ISTAT platform~\cite{istat_iscrizioni}.
Each record represents an individual who has formally deregistered from an Italian
municipal registry and declared a foreign country as their new place of residence.

This measurement mechanism is structurally tied to a human decision: a person
must actively initiate the deregistration procedure, either in person at their
home municipality or, since 2012, through the AIRE online portal.
As a result, the ISTAT series measures not emigration itself, but the
\emph{administratively visible} fraction of emigration those individuals
who chose to make their departure official.

For this study, we filter the dataset to retain only records with Italian
citizenship, total sex, all
age classes, covering the years from 2018 to 2022 and the
eight destination countries for which a corresponding Eurostat observation is
available.

\subsection{Eurostat: Immigration Flows from the Destination Side}
\label{subsec:data_eurostat}

The destination-country perspective is drawn from the Eurostat dataset
\texttt{MIGR\_IMM1CTZ} (\textit{Immigration by age group, sex and
citizenship})~\cite{eurostat_migr_imm1ctz}.
This dataset aggregates immigration records submitted by each member state to
Eurostat under Regulation (EC) No.~862/2007~\cite{ec_regulation_862_2007},
which standardises the definition of a long-term migrant as a person
establishing residence for at least twelve months.

For each country-year pair, we extract the total count of Italian citizens recorded as new immigrants,
using the completed-years age definition, all ages,
and both sexes combined.
The dataset covers eight destination countries: Austria, Denmark, Finland,
France, Germany, Norway, Spain, and Sweden, for the period from 2018 to 2022.
Two initially candidate countries (Poland and Portugal) returned empty
observations for all years in this dataset, indicating that those national
statistical offices did not submit disaggregated citizenship data to Eurostat
for the reference period.
This absence is itself informative: it reflects the uneven quality of migration
reporting across European systems.

\subsection{Dataset Construction and Alignment}
\label{subsec:data_alignment}

The two series are aligned on the (country, year) key, producing a panel of
40~observations across 8~countries and 5~years (2018--2022).
For each cell, we compute the raw gap as the difference between the Eurostat
count and the ISTAT count.

Two definitional asymmetries deserve explicit acknowledgment.
First, ISTAT records the year of deregistration, while Eurostat records the
year of arrival; for individuals who deregister with a delay, the two figures
may fall in different calendar years, introducing a phase offset.
Second, age disaggregation is not consistently available across destination
countries in \texttt{MIGR\_IMM1CTZ}; consequently, the analysis covers the
entire emigrant population rather than a specific age cohort.

\paragraph{Ensuring population comparability.}
A non-trivial alignment problem concerns the scope of the two series.
The ISTAT cancellation registry records \emph{all} individuals who
deregister from an Italian municipality, regardless of citizenship:
this includes foreign nationals previously resident in Italy who
relocate abroad.
Left unfiltered, the ISTAT figures would overcount Italian emigration
by conflating two structurally different phenomena: the departure of
Italian citizens and the onward migration of foreign residents.
We address this by restricting the ISTAT extract to records with
Italian citizenship, retaining only those individuals for whom the AIRE registration decision is actually
relevant.
On the Eurostat side, the \texttt{MIGR\_IMM1CTZ} dataset is natively filtered by the citizenship of the immigrant,
so no additional transformation is required.
After this filtering step, both series measure the same underlying
population: Italian citizens crossing an international border and
establishing residence abroad.